\documentclass[../main.tex]{subfiles}

\begin{document}

\chapter*{About The AI Engineer's Library}
\addcontentsline{toc}{chapter}{About The AI Engineer's Library}

\textit{The AI Engineer's Library} is a comprehensive five-volume series designed to take you from mathematical foundations to the cutting edge of artificial intelligence. Each book builds on the previous, creating a complete curriculum for the modern AI practitioner.

\vspace{1cm}

\section*{The Series}

\subsection*{Book 1: The Math That Powers AI}

The essential mathematical toolkit: linear algebra, probability, information theory, calculus, and numerical methods. This volume builds the foundation that everything else rests upon.

\vspace{0.5cm}

\subsection*{Book 2: Optimization for AI (This Book)}

From gradient descent to Adam, from convex optimization to the loss landscapes of deep networks. This volume covers the algorithms that train neural networks and the theory that explains when and why they work.

\vspace{0.5cm}

\subsection*{Book 3: The Generative Revolution}

The architectures that changed everything: transformers, attention mechanisms, diffusion models, and the generative AI systems reshaping our world. Understand how GPT, DALL-E, and their successors actually work.

\vspace{0.5cm}

\subsection*{Book 4: Inside the Black Box}

Interpretability, alignment, and the quest to understand what neural networks really learn. Essential reading for anyone building AI systems that need to be trustworthy, safe, and explainable.

\vspace{0.5cm}

\subsection*{Book 5: The Age of AI Agents}

Autonomous AI systems that can reason, plan, and act in the world. Tool use, multi-agent systems, and the emerging paradigm of AI that doesn't just answer questions but accomplishes goals.

\vspace{1cm}

\begin{center}
\fbox{\parbox{0.85\textwidth}{
\centering
\textbf{The Journey}\\[0.5em]
Together, these five volumes provide a complete path from mathematical novice to AI expert. Whether you read them in sequence or jump to the topics most relevant to your work, \textit{The AI Engineer's Library} will be your guide to mastering the field.
}}
\end{center}

\cleardoublepage
\end{document}
