\documentclass[../main.tex]{subfiles}
\begin{document}

\chapter*{Acknowledgments}
\addcontentsline{toc}{chapter}{Acknowledgments}
\thispagestyle{empty}

Writing a book that bridges mathematical rigor with practical AI applications requires standing on the shoulders of giants.

I am deeply grateful to the educators whose textbooks shaped my own understanding: Gilbert Strang for making linear algebra visual and intuitive, Stephen Boyd for the definitive treatment of convex optimization, and Christopher Bishop for connecting probability to machine learning so elegantly.

To the early readers who provided invaluable feedback on drafts---your questions revealed where my explanations needed work, and your enthusiasm kept me motivated through the long nights of revision.

To the open-source community, whose tools made this book possible: the NumPy, PyTorch, and JAX teams who built the foundations; the Jupyter project for enabling reproducible exploration; and the countless contributors to the libraries and datasets referenced throughout.

To my students over the years, whose confusion taught me what clarity means and whose questions often led to better explanations than I could have devised alone.

To my family, for their patience during the months when ``I'm just finishing this chapter'' became a recurring theme at dinner.

And to you, the reader, for trusting me to guide you through this material. I hope this book serves you as well as the books that shaped my own journey.

\vspace{1cm}
\hfill\textit{Dr. Vijay Raghavan}\\
\hfill\textit{April 2026}

\cleardoublepage
\end{document}
